---
title: "ST558 Homework 2"
format: html
editor: source
---

# Chris's Version

## Task 1: Basic Vector Practice
We start by creating two atomic vectors for our sample data and naming them via the given scheme:
```{r}
# create vectors
pre <- c(130, 128, 116, 124, 133, 134, 118, 126, 114, 127, 141, 138, 128, 140, 137, 131, 120, 128, 139, 135)
post <- c(114, 98, 113, 99, 107, 116, 113, 111, 119, 117, 101, 119, 130, 122, 106, 106, 124, 102, 117, 113)

# name vectors
names(pre) <- paste("Subject", 1:20, sep = "_")
names(post) <- paste("Subject", 1:20, sep = "_")

# preview
head(pre)
```

Now that we have our vectors, we can start performing operations on them:
```{r}
# create delta vector
delta <- pre - post

head(delta)

# find mean delta
mean_delta <- mean(delta)

mean_delta
```
The patients in our sample experienced a mean decrease of `r mean_delta` units during the study

Next, we'll use the `which()` function to determine which patients saw a positive decrease in blood pressure. I store these values in a vector called `pos_delta_idx` so that we can subset the original `delta` vector for the following question
```{r}
pos_delta_idx <- which(delta > 0)

pos_delta_idx

pos_delta <- delta[pos_delta_idx]

pos_delta

```
Lastly, we can determine the mean decrease among the patients who experienced a positive decrease:
```{r}
mean_pos_delta <- mean(pos_delta)

mean_pos_delta
```
Patients who experienced a positive decrease had a mean decrease of `r round(mean_pos_delta, 3)` units.

## Task 2: Basic Data Frame Practice
Now we're using data frames. I'm having to re-learn base R functions because I'm used to dplyr haha
```{r}
# create patient data frame
patient_df <- data.frame(patient = names(pre),
                         pre_bp = pre,
                         post_bp = post,
                         diff = delta,
                         row.names = NULL # removes vestigial "subject_x" row names 
                        )

# preview
head(patient_df)
```
Sweet, we have our data frame. I removed the row names inherited from our vectors because the data frame already has an index column. Now we can subset it and see which patients had negative differences in blood pressure during the experiment:
```{r}
# subset df
neg_diff_bp <- patient_df[patient_df$diff < 0,]

neg_diff_bp
```
*Subjects 9, 13, and 17* experienced a negative difference in blood pressure.

Next, we'll make an indicator variable to flag patients with a post-experiment BP of less than 120.
```{r}
# add low_post_bp indicator using vectorized ifelse (when do we learn mutate!!)
patient_df$low_post_bp <- ifelse(patient_df$post_bp < 120, TRUE, FALSE)

head(patient_df)
```
It looks like our indicator flag worked. Lastly we'll print the table:
```{r}
# print table with given code
knitr::kable(patient_df)
```
