---
title: "ST558 Homework 2"
format:
  html:
    toc: true
    theme: cosmo
editor: source
---

# Chris's Version

## Task 1: Basic Vector Practice
We start by creating two atomic vectors for our sample data and naming them via the given scheme:
```{r}
# create vectors
pre <- c(130, 128, 116, 124, 133, 134, 118, 126, 114, 127, 141, 138, 128, 140, 137, 131, 120, 128, 139, 135)
post <- c(114, 98, 113, 99, 107, 116, 113, 111, 119, 117, 101, 119, 130, 122, 106, 106, 124, 102, 117, 113)

# name vectors
names(pre) <- paste("Subject", 1:20, sep = "_")
names(post) <- paste("Subject", 1:20, sep = "_")

# preview
head(pre)
```

Now that we have our vectors, we can start performing operations on them:
```{r}
# create delta vector
delta <- pre - post

head(delta)

# find mean delta
mean_delta <- mean(delta)

mean_delta
```
The patients in our sample experienced a mean decrease of `r mean_delta` units during the study

Next, we'll use the `which()` function to determine which patients saw a positive decrease in blood pressure. I store these values in a vector called `pos_delta_idx` so that we can subset the original `delta` vector for the following question
```{r}
pos_delta_idx <- which(delta > 0)

pos_delta_idx

pos_delta <- delta[pos_delta_idx]

pos_delta

```
Lastly, we can determine the mean decrease among the patients who experienced a positive decrease:
```{r}
mean_pos_delta <- mean(pos_delta)

mean_pos_delta
```
Patients who experienced a positive decrease had a mean decrease of `r round(mean_pos_delta, 3)` units.

## Task 2: Basic Data Frame Practice
Now we're using data frames. I'm having to re-learn base R functions because I'm used to dplyr haha
```{r}
# create patient data frame
patient_df <- data.frame(patient = names(pre),
                         pre_bp = pre,
                         post_bp = post,
                         diff = delta,
                         row.names = NULL # removes goofy "subject_x" row names 
                        )

# preview
head(patient_df)
```
Sweet, we have our data frame. I removed the row names inherited from our vectors because the data frame already has an index column. Now we can subset it and see which patients had negative differences in blood pressure during the experiment:
```{r}
# subset df
neg_diff_bp <- patient_df[patient_df$diff < 0,]

neg_diff_bp
```
**Subjects 9, 13, and 17** experienced a negative difference in blood pressure.

Next, we'll make an indicator variable to flag patients with a post-experiment BP of less than 120.
```{r}
# add low_post_bp indicator using vectorized ifelse (when do we learn mutate!!)
patient_df$low_post_bp <- ifelse(patient_df$post_bp < 120, TRUE, FALSE)

head(patient_df)
```
It looks like our indicator flag worked. Lastly we'll print the table:
```{r}
# print table with given code
knitr::kable(patient_df)
```

## Task 3: List Practice
For this part, we start by making a placebo data frame with the same structure as the previous.
```{r}
# treatment data
placebo_pre <- c(138, 135, 147, 117, 152, 134, 114, 121, 131, 130)
placebo_post <- c(105, 136, 123, 130, 134, 143, 135, 139, 120, 124)

# make df with all the bells and whistles
placebo_df <- data.frame(patient = paste("Subject", 1:length(placebo_pre), sep = "_"),
                         pre_bp = placebo_pre,
                         post_bp = placebo_post,
                         diff = placebo_pre - placebo_post,
                         low_post_bp = ifelse(placebo_post < 120, TRUE, FALSE),
                         row.names = NULL # removes vestigial "subject_x" row names
                        )
               
# preview
head(placebo_df)
```
Following part 2, we create a list with both data frames and access the elements a few different ways:
```{r}
# make list
treatment_list <- list(treatment = patient_df,
                       placebo = placebo_df)

# accessing first list 3 ways:
treatment_list[1] # basic bracket
treatment_list[[1]] # simplified
treatment_list$treatment # using $

# accessing the pre_bp column from the placebo DF
treatment_list$placebo[,"post_bp", drop = FALSE]
```
Very cool.

## Task 4: Control Loop Practice
We're now using a `for` loop to iterate through each data frame and create levels of a new variable called `status`. I'm choosing to make `status` a factor instead of a character variable since we have discrete, unchanging levels.
```{r}
# create status column
treatment_list$treatment$status <- factor(character(nrow(treatment_list$treatment)),
                                          levels = (c("Optimal", "Borderline", "High")),
                                          ordered = TRUE)
treatment_list$placebo$status <- factor(character(nrow(treatment_list$placebo)),
                                        levels = (c("Optimal", "Borderline", "High")),
                                        ordered = TRUE)

# assign levels of status to each obs in treatment df
for(i in 1:nrow(treatment_list$treatment)) {
  if(treatment_list$treatment$post_bp[i] > 130) {
    treatment_list$treatment$status[i] <- "High"
  } else if(treatment_list$treatment$post_bp[i] >= 120) {
    treatment_list$treatment$status[i] <- "Borderline"
  } else {
    treatment_list$treatment$status[i] <- "Optimal"
  }
}

# preview
head(treatment_list$treatment[,-2])
```
Looks good, we can now repeat the same thing for the placebo data frame:
```{r}
# assign levels of status to each obs in placebo df
for(i in 1:nrow(treatment_list$placebo)) {
  if(treatment_list$placebo$post_bp[i] > 130) {
    treatment_list$placebo$status[i] <- "High"
  } else if(treatment_list$placebo$post_bp[i] >= 120) {
    treatment_list$placebo$status[i] <- "Borderline"
  } else {
    treatment_list$placebo$status[i] <- "Optimal"
  }
}

# preview
head(treatment_list$placebo[,-2])
```
Beautiful!

## Task 5: Function Writing
Originally I used a for loop to traverse each element in the `stat_vec` vector but then I realized `lapply()` and `sapply()` are base R. Much more succinct, not that statisticians care about silly things like runtime.
```{r}
find.stats <- function(treatment_list, stat = "mean") {
  # pull relevant columns from treatment/placebo
  treatment_data <- treatment_list$treatment[c("pre_bp", "post_bp", "diff")]
  placebo_data <- treatment_list$placebo[c("pre_bp", "post_bp", "diff")]
  
  # concatenate data, make result vector
  stat_vec <- c(treatment_data, placebo_data)
  result_vec <- sapply(stat_vec, get(stat))

  # make name vector using paste
  names_vec <- paste(rep(c("treatment", "placebo"), each = 3),
                     rep(c("pre_bp", "post_bp", "diff"), 2),
                     rep(stat, 6),
                     sep = "_")
  
  # name result vector then return it
  names(result_vec) <- names_vec
  return(result_vec)
}
```

Before we test out the function, I'm calling `summary()` to confirm that the values are accurate:
```{r}
# top output is treatment, bottom is placebo
summary(patient_df[c("pre_bp", "post_bp", "diff")])
summary(placebo_df[c("pre_bp", "post_bp", "diff")])
```

Now to try out our function (without specifying the stat):
```{r}
find.stats(treatment_list)
```
It's alive! The values are correct so we can see how it does with other statistics:
```{r}
find.stats(treatment_list, "var")
find.stats(treatment_list, "sd")
find.stats(treatment_list, "min")
find.stats(treatment_list, "max")
```
A wall of stats and no errors. Fantastic.