---
title: "ST558 Homework 2"
format: html
editor: source
---

# Chris's Version

# Task 1: Basic Vector Practice
We start by creating two atomic vectors for our sample data and naming them via the given scheme:
```{r}
# create vectors
pre <- c(130, 128, 116, 124, 133, 134, 118, 126, 114, 127, 141, 138, 128, 140, 137, 131, 120, 128, 139, 135)
post <- c(114, 98, 113, 99, 107, 116, 113, 111, 119, 117, 101, 119, 130, 122, 106, 106, 124, 102, 117, 113)

# name vectors
names(pre) <- paste("Subject", 1:20, sep = "_")
names(post) <- paste("Subject", 1:20, sep = "_")

# preview
head(pre)
```

Now that we have our vectors, we can start performing operations on them:
```{r}
# create delta vector
delta <- pre - post

head(delta)

# find mean delta
mean_delta <- mean(delta)

mean_delta
```
The patients in our sample experienced a mean decrease of `r mean_delta` units during the study

Next, we'll use the `which()` function to determine which patients saw a positive decrease in blood pressure. I store these values in a vector called `pos_delta_idx` so that we can subset the original `delta` vector for the following question
```{r}
pos_delta_idx <- which(delta > 0)

pos_delta_idx

pos_delta <- delta[pos_delta_idx]

pos_delta

```
Lastly, we can determine the mean decrease among the patients who experienced a positive decrease:
```{r}
mean_pos_delta <- mean(pos_delta)

mean_pos_delta
```
Patients who experienced a positive decrease had a mean decrease of `r round(mean_pos_delta, 3)` units.

